Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-04-25 WSSSPE2_report/WSSSPE2_review.tex
Section 6: minor changes. Other sections: change a double quote to two single quotes.

\subsubsection{Subgroup B Discussion}
The statement ``applied computer science is being attempted in academia
without any formal training'' kicked off our discussion. The group brought
expertise in several different training models, from a 2-day Software
Carpentry workshop, to a 3-week Open Science for Synthesis (OSS) program
to semester-long programs. We discussed the pros and cons of different
training approaches, also touching on informal learning, for example where
people learning the necessary skills because they had cross-disciplinary
people (``boundary scientists'') in their work environments of whom they
could ask questions.

Some papers explored gaps in training of early career scientists. Industry
participants in our group confirm this. We asked ourselves, ``Are
traditional courses failing?'' We think yes. Changes to undergraduate
curricula requirements are difficult. But as programming models become
more complex, we have to raise the skill levels. Skills must be improved
not only in traditional programming - learning languages and algorithms -
but around professional software development. We want to get to a point
where ``Everyone has a new minimum now - everyone knows git.'' Developers
also need training on licensing choices. Just because a code is on git
doesn't mean it's open.

"How do you get people to look for the training they need?'' wondered one
participant. People seek out opportunities like Software Carpentry to
augment skills. While some instruction can be done in an institutional
curricula, independent groups (non-profits, institutes) have more
flexibility. Some asked whether Silicon Valley be interested in funding
training so that people are better prepared to enter the workforce? Some
felt that companies were reluctant to deliver training because employees
then leave the company. While others felt that pushing all training to
industry could lead to good technical people ``getting on the Google bus''
and leaving the sciences.

"How will we know when people are trained effectively in these new
skills?" We discussed certification. It can be hard to build the
recognition of Java, database certifications amongst all technologies. OSS
offers badges to those completing training, however we need to demonstrate
proficiency, not just completion. Google Summer of Code is a big CV
augmentation for participants, can we duplicate something like that?

The group also discussed how to fund training. On participant observed
that NYU runs a 6-week data science training, companies grab graduates but
pay for those they hire. OSS also used the NSF Software Institutes program
as a vehicle to fund training activities. Software Carpentry uses a
collaborative teaching approach where  people publish open teaching
materials and receive credit for their use.

@@

We then discussed variations amongst scientific domain areas and wondered
"Are some communities more or less open than others?'' To some members of
the group biology seems to be open, while physics less so. Some felt with
the more recent development of bioinformatics as a field there were fewer
historical practices to undo. Physics has pre-print philosophy to
overcome. Environmental sciences may be mixed. The group felt the
biomedical area, however, was very competitive and closed. One participant
mentioned blueprints for going open source (like NWChem recently did)
where authors outline how this helps, what you do and what the next steps
are.

We then moved beyond our training discussion to address funding that
encourages sustainable software -- funding of both people and of projects
that create a true system to support sustainability. We called this
``institutionalized serendipity.'' Because science is increasingly reliant
on software, one participant observed that ``software development can have
much broader impact than publishing individual research, but it is not
viewed that way.'' Because of this centrality, one participant mentioned
that training ought to be called Science Carpentry rather than Software
Carpentry. We felt we were beginning to see changes in the research
community as a result of the NSF's data management plan requirements. PIs
are thinking more about data and some university libraries are offering
data repositories. We wondered if a software management plan might be
effective. "How might funding programs need to adapt to reward good
software development practices?'' we asked ourselves. We thought about
best practices such as version control, test harnesses, mailing lists, bug
tracking, community contribution, and reuse where appropriate. We thought
about measuring success through usage statistics (downloads, altmetrics).
"Should funders demand reproducibility?'' we wondered. In order for results
to be reproduced, software would need to be carefully curated.

Again, our industry participants contributed unique viewpoints. Partnering
with industry was seen one path to sustainability. The unique partnerships
Kitware engages in promote academic freedom while creating an income
stream from certain portions of the software. This type of approach to
sustainable software frees researchers from performing tasks that do not
offer the rewards their institution values.'' We talked some about
successful models for industry partnerships that preserve open science.
Participants noted that there are some NSF programs that prohibit
partnerships with for profit companies (but there are others where this is
encouraged).

@@

Subgroup B then focused on actions it could take. It discussed development
of a white paper that describes a matrixed approach to training
(multi-day, multi-week, semester). The white paper might include a survey
of existing techniques. There are many, some dating back many years, for
example the Interuniversity Consortium for Political and Social Research
(ICPSR) and various summer institutes. The white paper could Include a
call for a comprehensive assessment of these techniques. We need to think
carefully about the right venue for such a white paper, where would it
have the most impact. The subgroup believes it would need to approach editors
directly to ascertain this.

The group felt that training in techniques that promote sustainability has
a range of benefits: career paths, educated reviewers, reproducible
science and more. There is some information on how that has been
approached and assessed, but more is needed. The group felt that this
training is undervalued and that it's important to communicate the return
on investment - both individual ROI (skills that make scientists more
effective and more marketable) and funder ROI (better use of taxpayer
funds, research more likely to be reproducible because sustainable
software exists, better trained reviewers).

@@

\begin{itemize}
\item Michael R. Crusoe and C.Titus Brown. Channeling community contributions to
scientific software: a hackathon experience~\cite{wssspe2_crusoe}

@@

This group started by discussing who should be incentivized, thinking of two
categories of people: those in science (who could be incentivized to do better,
more shareable, more sustainable work), and those in industry but interested in
science (who could be incentivized to contribute to science.) It was pointed out
that we are not yet clear enough on exactly what we want to incentivize,
suggesting that we need to have a clearer picture of ``good computational work"
and what sort of contributions are truly generative for science.

The discussion of incentivizing those in science acknowledged that the
publications system was far from perfect for incentivizing good software work.
Nonetheless discussion focused on bringing software people into publications.
There were two main suggestions. The first is to focus on end users of software
and encourage them to cite software the ``right" way. James Howison suggested
that his research showed that few projects were making a formal request for
citation (but that authors weren't necessarily following those suggestions
anyway)~\cite{howison2015jasist}. He suggested making access to the software
conditional on a license that requires citation. Others found this ``too
confrontational" and preferred to concentrate on making it easier to do the
right thing. The second was focusing on those leading software projects, and the
group was more enthusiastic about ``forcing" PIs to include their ``software
people" on publications, although there were few ideas on how exactly to do
this. Another technique mentioned was that when scientific software projects are
hosted in organizations like Apache, the scientific contributors can benefit
from building their reputations, perhaps yielding job offers that they can use
to negotiate better job and career packages.

@@

The group also discussed developing ``software prizes" arguing that while it is
hard to ``mint" other new sources of reputation, prizes are possible without
getting too many others on board. The prize criteria can form a template for
describing what we mean by scientific contributions made through software,
particularly focusing on building active communities, not only writing great
code.

>>>>

\subsubsection{Subgroup B Discussion}
The statement ``applied computer science is being attempted in academia
without any formal training'' kicked off our discussion. The group brought
expertise in several different training models, from a two-day Software
Carpentry workshop, to a three-week Open Science for Synthesis (OSS) program
to semester-long programs. We discussed the pros and cons of different
training approaches, also touching on informal learning, for example where
people learning the necessary skills because they had cross-disciplinary
people (``boundary scientists'') in their work environments of whom they
could ask questions.

Some papers explored gaps in training of early career scientists. Industry
participants in our group confirm this. We asked ourselves, ``Are
traditional courses failing?'' We think yes. Changes to undergraduate
curricula requirements are difficult. But as programming models become
more complex, we have to raise the skill levels. Skills must be improved
not only in traditional programming -- learning languages and algorithms --
but around professional software development. We want to get to a point
where ``Everyone has a new minimum now -- everyone knows Git.'' Developers
also need training on licensing choices. Just because a code is on Git
does not mean that it is open.

``How do you get people to look for the training they need?'' wondered one
participant. People seek out opportunities like Software Carpentry to
augment skills. While some instruction can be done in an institutional
curricula, independent groups (non-profits, institutes) have more
flexibility. Some asked whether Silicon Valley be interested in funding
training so that people are better prepared to enter the workforce? Some
felt that companies were reluctant to deliver training because employees
then leave the company. While others felt that pushing all training to
industry could lead to good technical people ``getting on the Google bus''
and leaving the sciences.

``How will we know when people are trained effectively in these new
skills?'' We discussed certification. It can be hard to build the
recognition of Java or database certifications amongst all technologies. OSS
offers badges to those completing training. However we need to demonstrate
proficiency, not just completion. Google Summer of Code is a big CV
augmentation for participants---can we duplicate something similar?

The group also discussed how to fund training. On participant observed
that NYU runs a six-week data-science training; companies grab the graduates and
pay for those they hire. OSS also used the NSF Software Institutes program
as a vehicle to fund training activities. Software Carpentry uses a
collaborative teaching approach where  people publish open teaching
materials and receive credit for their use.

@@

We then discussed variations amongst scientific domain areas and wondered
``Are some communities more or less open than others?'' To some members of
the group biology seems to be open, while physics less so. Some felt with
the more recent development of bioinformatics as a field there were fewer
historical practices to undo. Physics has pre-print philosophy to
overcome. Environmental sciences may be mixed. The group felt the
biomedical area, however, was very competitive and closed. One participant
mentioned blueprints for going open source (like NWChem recently did)
where authors outline how this helps, what you do and what the next steps
are.

We then moved beyond our training discussion to address funding that
encourages sustainable software -- funding of both people and of projects
that create a true system to support sustainability. We called this
``institutionalized serendipity.'' Because science is increasingly reliant
on software, one participant observed that ``software development can have
much broader impact than publishing individual research, but it is not
viewed that way.'' Because of this centrality, one participant mentioned
that training ought to be called Science Carpentry rather than Software
Carpentry. We felt we were beginning to see changes in the research
community as a result of the NSF's data management plan requirements. PIs
are thinking more about data and some university libraries are offering
data repositories. We wondered if a software management plan might be
effective. ``How might funding programs need to adapt to reward good
software development practices?'' we asked ourselves. We thought about
best practices such as version control, test harnesses, mailing lists, bug
tracking, community contribution, and reuse where appropriate. We thought
about measuring success through usage statistics (downloads, altmetrics).
``Should funders demand reproducibility?'' we wondered. In order for results
to be reproduced, software would need to be carefully curated.

Again, our industry participants contributed unique viewpoints. Partnering
with industry was seen as one path to sustainability. The unique partnerships
Kitware engages in promote academic freedom while creating an income
stream from certain portions of the software. This type of approach to
sustainable software frees researchers from performing tasks that do not
offer  the rewards their institutions value. We also discussed  about
successful models for industry partnerships that preserve open science.
Participants noted that there are some NSF programs that prohibit
partnerships with for-profit companies (but there are other programs in which this is
encouraged).

@@

Subgroup B then focused on actions it could take. It discussed development
of a white paper that describes a matrix approach to training
(multi-day, multi-week, semester). The white paper might include a survey
of existing techniques. There are many, some dating back many years, for
example the Interuniversity Consortium for Political and Social Research
(ICPSR) and various summer institutes. The white paper could Include a
call for a comprehensive assessment of these techniques. We need to think
carefully about the right venue for such a white paper, where would it
have the most impact. The subgroup believes it would need to approach editors
directly to ascertain this.

The group felt that training in techniques that promote sustainability has
a range of benefits: career paths, educated reviewers, reproducible
science and more. There is some information on how that has been
approached and assessed, but more is needed. The group felt that this
training is undervalued and that it is important to communicate the return
on investment -- both individual ROI (skills that make scientists more
effective and more marketable) and funder ROI (better use of taxpayer
funds, research more likely to be reproducible because sustainable
software exists, better trained reviewers).

@@

\begin{itemize}
\item Michael R. Crusoe and C. Titus Brown. Channeling community contributions to
scientific software: a hackathon experience~\cite{wssspe2_crusoe}

@@

This group started by discussing who should be incentivized, thinking of two
categories of people: those in science (who could be incentivized to do better,
more shareable, more sustainable work), and those in industry but interested in
science (who could be incentivized to contribute to science.) It was pointed out
that we are not yet clear enough on exactly what we want to incentivize,
suggesting that we need to have a clearer picture of ``good computational work''
and what sort of contributions are truly generative for science.

The discussion of incentivizing those in science acknowledged that the
publications system was far from perfect for incentivizing good software work.
Nonetheless discussion focused on bringing software people into publications.
There were two main suggestions. The first is to focus on end users of software
and encourage them to cite software the ``right'' way. James Howison suggested
that his research showed that few projects were making a formal request for
citation (but that authors weren't necessarily following those suggestions
anyway)~\cite{howison2015jasist}. He suggested making access to the software
conditional on a license that requires citation. Others found this ``too
confrontational'' and preferred to concentrate on making it easier to do the
right thing. The second was focusing on those leading software projects, and the
group was more enthusiastic about ``forcing'' PIs to include their ``software
people'' on publications, although there were few ideas on how exactly to do
this. Another technique mentioned was that when scientific software projects are
hosted in organizations like Apache, the scientific contributors can benefit
from building their reputations, perhaps yielding job offers that they can use
to negotiate better job and career packages.

@@

The group also discussed developing ``software prizes'' arguing that while it is
hard to ``mint'' other new sources of reputation, prizes are possible without
getting too many others on board. The prize criteria can form a template for
describing what we mean by scientific contributions made through software,
particularly focusing on building active communities, not only writing great
code.
